---
title: "Data Science Capstone Milestone Report"
author: "Matthew Sedlar"
date: "March 20, 2016"
output: html_document
---

This report contains an exploratory analysis of data used in the Coursera/Johns Hopkins Bloomberg School of Public Health Data Science Capstone Project. 

## The Data

This project is an attempt to create a predictive text model using text mining procedures. The data, supplied by Coursera in partnership with SwiftKey, consists of three large text documents split by various sources -- blogs, news reports and Twitter.

Basic summaries of the text documents follow:

```{r echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}

require(stringi)

# this is the method used to count lines and words below
# but is not integrated into the report because it's a
# memory intensive process

twitter.doc <- stri_conv(readLines("data/en_US.twitter.txt"),"UTF-8")
blog.doc <- stri_conv(readLines("data/en_US.blogs.txt"),"UTF-8")
news.doc <- stri_conv(readLines("data/en_US.news.txt"),"UTF-8")

print(paste("News, ", stri_stats_general(news.doc)[[1]],
           " lines of text, containing ",
           stri_stats_latex(news.doc)[[4]], " words.", sep = ""))
print(paste("Blogs, ", stri_stats_general(blog.doc)[[1]],
           " lines of text, containing ",
           stri_stats_latex(blog.doc)[[4]], " words.", sep = ""))
print(paste("Twitter, ", stri_stats_general(twitter.doc)[[1]],
           " lines of text, containing ",
           stri_stats_latex(twitter.doc)[[4]], " words.", sep = ""))

```

* News, 1,010,242 lines of text, containing 34,494,539 words.
* Blogs, 899,288 lines of text, containing 37,570,839 words.
* Twitter, 2,360,148 lines of text, containing 30,451,128 words.

Because of the size of each document, I chose to use the `sample_lines` function in the LaF package to sample 1.5% of lines from each document and create a corpus of the text.

```{r message=FALSE}

require(tm)
require(LaF)

# read lines randomly into a data frame
set.seed(1234)

docs <- c(data.frame(sample_lines("data/en_US.twitter.txt", 
                                  .015 * 2360148,
                                  2360148)),
          data.frame(sample_lines("data/en_US.blogs.txt",
                                  .015 * 899288,
                                  899288)),
          data.frame(sample_lines("data/en_US.news.txt",
                                  .015 * 1010242,
                                  1010242)))

c <- Corpus(VectorSource(docs))

```

## Cleaning the Data


```{r}

# eliminate whitespace
c <- tm_map(c, stripWhitespace)
# remove all punctuation except apostrophes, fix 'I'm'
(f <- content_transformer(function(x,pattern,sub) gsub(pattern,sub,x)))
c <- tm_map(c, f, "[^[:alnum:][:space:]']","")
c <- tm_map(c, f, "(i'm)","I'm")
# to lower case
c <- tm_map(c, content_transformer(tolower))
# remove numbers
c <- tm_map(c, content_transformer(removeNumbers))

# profanity filtering
bad.words <- c()
bad.nouns <- c("ass","nigg","dick")
bad.verbs <- c("fuck","shit")

bad.words <- c(bad.words,
               paste(bad.nouns,"(a|as|es|er?)?s?|",sep=""),
               paste(bad.verbs,"(a|e|ed|er|ing?)?s?|",sep=""))

bad.words <- paste(bad.words,collapse="")
c <- tm_map(c, f, bad.words,"")

```

Once the data is cleaned, I can turn it into a term document matrix and get a better idea of what it looks like. I'm interested in both 1-grams and 2-grams, or frequently occuring words and two-word combinations. Below are two visualizations of the data: a histogram for 1-gram and 2-grams as well as wordclouds.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# TOKENIZER
BigramTokenizer <- function(x) {
  unlist(lapply(ngrams(words(x),2),paste, collapse=" "), use.names = F)
}

# term document matrix, 1-gram and 2-gram
tdm.1g <- TermDocumentMatrix(c)
# tdm.2g <- TermDocumentMatrix(c, control=list(tokenize=BigramTokenizer))

tdm.1g.matrix <- as.matrix(tdm.1g)
#tdm.2g.matrix <- as.matrix(tmd.2g)

frequency.1g <- rowSums(tdm.1g.matrix)
#frequncey.2g <- rowSums(tmd.2g.matrix)

frequency.1g <- sort(frequency.1g, decreasing = T)
#frequency.2g <- sort(frequency.2g, decreasing = T)

```

## Has the data scientist made basic plots, such as histograms to illustrate features of the data?

```{r}

df <- data.frame(word=names(frequency.1g),freq=frequency.1g)
df$word <- factor(df$word, levels = df$word[order(-df$freq)])

require(ggplot2)

ggplot(df[1:25,]) + geom_bar(aes(word,freq,fill=freq),stat="identity") +
  ggtitle("Frequency of Top 25 1-grams") +
  ylab("Frequency") +
  xlab("Words") +
  theme(axis.text.x=element_text(angle=90),
        plot.title=element_text(size=15, face="bold"),
        legend.position="none")

```

## Was the report written in a brief, concise style, in a way that a non-data scientist manager could appreciate?
